---
title: "Introduction to xform_function"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to xform_function}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

This vignette provides examples of how to use the `xform_function` transformation to create new data features for PMML models.

Given a `xform_wrap` object and a transformation expression, `xform_function` calculates data for a new feature and creates a new `xform_wrap` object. When PMML is produced with `pmml::pmml()`, the transformation is inserted into the `LocalTransformations` node as a `DerivedField`.

`xform_function` makes it possible to use multiple data fields and functions to produce a new feature.

The `kable` function is part of `knitr`, and is used to make tables more readable.

```{r, echo=FALSE,warning=FALSE,message=FALSE,results="hide"}
library(pmml)
library(knitr)
```

## Single numeric field
Using the `iris` dataset as an example, let's construct a new feature by transforming one variable. Load the dataset and show the first few lines:
```{r}
data(iris)
kable(head(iris,3))
```

Create the `iris_box` object with `xform_wrap`:
```{r}
iris_box <- xform_wrap(iris)
```

`iris_box` contains the data and transform information that will be used to produce PMML later.
The original data is in `iris_box$data`. Any new features created with a transformation are added as columns to this data frame.
```{r}
kable(head(iris_box$data,3))
```

Transform and field information is in `iris_box$field_data`. The field_data data frame contains 
information on every field in the dataset, as well as every transform used. The `xform_function` column contains expressions used in 
the `xform_function` transform.
```{r}
kable(iris_box$field_data)
```

Now add a new feature, `Sepal.Length.Sqrt`, using `xform_function`:
```{r}
iris_box <- xform_function(iris_box,orig_field_name="Sepal.Length",
                         new_field_name="Sepal.Length.Sqrt",
                         expression="sqrt(Sepal.Length)")
```

The new feature is calculated and added as a column to the `iris_box$data` data frame:
```{r}
kable(head(iris_box$data,3))
```

`iris_box$field_data` now contains a new row with the transformation expression:
```{r}
kable(iris_box$field_data[6,c(1:3,14)])
```


Construct a linear model for `Petal.Width` using this new feature, and convert it to PMML:
```{r}
fit <- lm(Petal.Width ~ Sepal.Length.Sqrt, data=iris_box$data)
fit_pmml <- pmml(fit, transform=iris_box)
```

Since the model predicts `Petal.Width` using a variable based on `Sepal.Length`, the PMML will contain 
these two fields in the `DataDictionary` and `MiningSchema`:
```{r}
fit_pmml[[2]] #Data Dictionary node
fit_pmml[[3]][[1]] #Mining Schema node
```

The `LocalTransformations` node contains `Sepal.Length.Sqrt` as a derived field:
```{r}
fit_pmml[[3]][[3]]
```

## Single categorical field
`xform_function` can also operate on categorical data. In this example, let's create a boolean feature that equals 1 only when `Species` is `setosa`:
```{r}
iris_box <- xform_wrap(iris)
iris_box <- xform_function(iris_box,orig_field_name="Species",
                         new_field_name="Species.Setosa",
                         expression="if (Species == 'setosa') {1} else {0}")
kable(head(iris_box$data,3))
```

Create a linear model and check the `LocalTransformations` node:
```{r}
fit <- lm(Petal.Width ~ Species.Setosa, data=iris_box$data)
fit_pmml <- pmml(fit, transform=iris_box)
fit_pmml[[3]][[3]]
```

## Multiple input fields

It is possible to create new features by combining several fields. Let's create a new field from the ratio of sepal and petal lengths:
```{r}
iris_box <- xform_wrap(iris)
iris_box <- xform_function(iris_box,orig_field_name="Sepal.Length,Petal.Length",
                         new_field_name="Length.Ratio",
                         expression="Sepal.Length / Petal.Length")
```

As before, the new field is added as a column to the `iris_box$data` data frame:
```{r}
kable(head(iris_box$data,3))
```

Fit a linear model using this new feature, and convert it to pmml:
```{r}
fit <- lm(Petal.Width ~ Length.Ratio, data=iris_box$data)
fit_pmml <- pmml(fit, transform=iris_box)
```

The pmml will contain `Sepal.Length` and `Petal.Length` in the `DataDictionary` and `MiningSchema`, since these were used in `FormulaXform`:
```{r}
fit_pmml[[2]] #Data Dictionary node
fit_pmml[[3]][[1]] #Mining Schema node
```

The `Local.Transformations` node contains `Length.Ratio` as a derived field:
```{r}
fit_pmml[[3]][[3]]
```

## Using a previously derived feature
It is possible to pass a feature derived with `xform_function` to another `xform_function` call. To do this, the second call to `xform_function` must use the original data field names (instead of the derived field) in the `orig_field_name` argument.

```{r}
iris_box <- xform_wrap(iris)
iris_box <- xform_function(iris_box,orig_field_name="Sepal.Length,Petal.Length",
                         new_field_name="Length.Ratio",
                         expression="Sepal.Length / Petal.Length")

iris_box <- xform_function(iris_box,orig_field_name="Sepal.Length,Petal.Length,Sepal.Width",
                         new_field_name="Length.R.Times.S.Width",
                         expression="Length.Ratio * Sepal.Width")
kable(iris_box$field_data[6:7,c(1:3,14)])
```

```{r}
fit <- lm(Petal.Width ~ Length.R.Times.S.Width, data=iris_box$data)
fit_pmml <- pmml(fit, transform=iris_box)

```

The pmml will contain `Sepal.Length`, `Petal.Length`, and `Sepal.Width` in the `DataDictionary` and `MiningSchema`, since these were used in `FormulaXform`:
```{r}
fit_pmml[[2]] #Data Dictionary node
fit_pmml[[3]][[1]] #Mining Schema node
```

The `Local.Transformations` node contains `Length.Ratio` and `Length.R.Times.S.Width` as derived fields:
```{r}
fit_pmml[[3]][[3]]
```


## PMML functions supported by `xform_function`

The following R functions and operators are directly supported by `xform_function`. Their PMML equivalents are listed in the second column:
```{r,echo=FALSE}
R <- c("+","-","/","*","^","<","<=",">",">=","&&","&","|","||","==","!=","!","ceiling","prod","log")
PMML <- c("+","-","/","*","pow","lessThan","lessOrEqual","greaterThan","greaterOrEqual","and","and","or","or","equal","notEqual","not","ceil","product","ln")

funcs_df <- data.frame(R, PMML)
knitr::kable(funcs_df)
```



For these functions, no extra code is required for translation.

The R function `prod` can be used as long as only numeric arguments are specified. That is, `prod` can take an `na.rm` argument, but specifying this in `xform_function` directly will not produce PMML equivalent to the R expression.

Similarly, the R function `log` can be used directly as long as the second argument (the base) is not specified.


## PMML functions not supported by `xform_function`

There are built-in functions defined in PMML that cannot be directly translated to PMML using `xform_function` as described above.

In this case, an error will be thrown when R tries to calculate a new feature using the function passed to `xform_function`, but does not see that function in the environment.

It is still possible to make `xform_function` work, but the PMML function must be defined in the R environment first.

Let's use `isIn`, a PMML function, as an example. The function returns a boolean indicating whether the first argument is contained in a list of values. Detailed specification for this function is available on [this DMG page](http://dmg.org/pmml/v4-4-1/BuiltinFunctions.html#boolean5). 

One way to implement this in R is by using `%in%`, with the list of values being represented by `...`:
```{r}
isIn <- function(x, ...) {
  dots <- c(...)
  if (x %in% dots) {
    return(TRUE)
  } else {
    return(FALSE)
  }
}

isIn(1,2,1,4)
```

This function can now be passed to `xform_function`. The following code creates a feature that indicates whether `Species` is 
either `setosa` or `versicolor`:
```{r}
iris_box <- xform_wrap(iris)
iris_box <- xform_function(iris_box,orig_field_name="Species",
                         new_field_name="Species.Setosa.or.Versicolor",
                         expression="isIn(Species,'setosa','versicolor')")
```

The `data` data frame now contains the new feature:
```{r}
kable(head(iris_box$data,3))
```

Create a linear model and view the corresponding PMML for the function:
```{r}
fit <- lm(Petal.Width ~ Species.Setosa.or.Versicolor, data=iris_box$data)
fit_pmml <- pmml(fit, transform=iris_box)
fit_pmml[[3]][[3]]
```

## PMML Function not supported by `xform_function` - another example
As another example, let's use R's `mean` function to create a new feature. PMML has a built-in `avg`, so we will define an R function with this name.
```{r}
avg <- function(...) {
  dots <- c(...)
  return(mean(dots))
}
```
Now use this function to take an average of several other features and combine with another field:
```{r}
iris_box <- xform_wrap(iris)
iris_box <- xform_function(iris_box,orig_field_name="Sepal.Length,Petal.Length,Sepal.Width",
                         new_field_name="Length.Average.Ratio",
                         expression="avg(Sepal.Length,Petal.Length)/Sepal.Width")
```
The `data` data frame now contains the new feature:
```{r}
kable(head(iris_box$data,3))
```

Create a simple linear model and view the corresponding PMML for the function:
```{r}
fit <- lm(Petal.Width ~ Length.Average.Ratio, data=iris_box$data)
fit_pmml <- pmml(fit, transform=iris_box)
fit_pmml[[3]][[3]]
```

In the PMML, `avg` will be recognized as a valid function.



## PMML for arbitrary functions
The function `function_to_pmml` (part of the `pmml` package) makes it possible to convert an R expression into PMML directly, without creating a model or calculating values.

As long as the expression passed to the function is a valid R expression (e.g., no unbalanced parentheses), it can contain arbitrary function names not defined in R. Variables in the expression passed to `xform_function` are always assumed to be field names, and not substituted. That is, even if `x` has a value in the R environment, the resulting expression will still use `x`.

```{r}
function_to_pmml("1 + 2")

x <- 3
function_to_pmml("foo(bar(x * y))")
```


## More notes on functions
There are several limitations to parsing expressions in `xform_function`.

Each transformation operates on one data row at a time. For example, it is not possible to compute the mean of an entire feature column in `xform_function`.

An expression such as `foo(x)` is treated as a function `foo` with argument `x`. Consequently, passing in an R vector `c(1,2,3)` will produce PMML where `c` is a function and `1,2,3` are the arguments:
```{r}
function_to_pmml("c(1,2,3)")
```

We can also see what happens when passing an `na.rm` argument to `prod`, as mentioned in an above example:
```{r}
function_to_pmml("prod(1,2,na.rm=FALSE)") #produces incorrect PMML
function_to_pmml("prod(1,2)") #produces correct PMML
```
Additionally, passing in a vector to `prod` produces incorrect PMML:
```{r}
prod(c(1,2,3))
function_to_pmml("prod(c(1,2,3))")
```


## More examples of functions
The following are additional examples of pmml produced from R expressions.

Extra parentheses:
```{r}
function_to_pmml("pmmlT(((1+2))*(x))")
```

If-else expressions:
```{r}
function_to_pmml("if(a<2) {x+3} else if (a>4) {4} else {5}")
```


## References
- [DMG PMML 4.4 specification](http://dmg.org/pmml/v4-4-1/GeneralStructure.html)